Avoiding tipping points in the management of ecological systems: a non-parametric Bayesian approach

### 2013-05-03 inst/doc/nonparametric-bayes.tex
draft for ERE

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{color}
\usepackage{fancyvrb}
\DefineShortVerb[commandchars=\\\{\}]{\|}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\newenvironment{Shaded}{}{}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.56,0.13,0.00}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{{#1}}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textit{{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{{#1}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{{#1}}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.02,0.16,0.49}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{{#1}}}}
\newcommand{\NormalTok}[1]{{#1}}
\usepackage{longtable}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Non-parametric approaches to optimal policy are more robust},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

\begin{document}
\begin{frontmatter}
  \title{Non-parametric approaches to optimal policy are more robust}
  \author[cstar]{Carl Boettiger\corref{cor1}}
  \cortext[cor1]{Corresponding author}
  \ead{cboettig@ucsc.edu}
  \author[cstar]{Marc Mangel}
  \author[noaa]{Stephan B. Munch}
  \address[cstar]{Center for Stock Assessment Research, Department of Applied Math and Statistics, University of California, Mail Stop SOE-2, Santa Cruz, CA 95064, USA}
  \address[noaa]{Southwest Fisheries Science Center, National Oceanic and Atmospheric Administration, 110 Shaffer Road, Santa Cruz, CA 95060, USA}
 \end{frontmatter}

@@

Most management recommendations from the ecological literature are based
on (or at least motivated by) parametric models. Though in principle
these models reflect mechanistic underpinnings of biological
interactions involved, in practice real mechanisms are often unknown and
the true dynamics too complex to be captured by simple models
{[}@refs{]}. While simple mechanistic models can nevertheless provide
important insights into possible dynamics -- for instance, demonstrating
that a vaccine does not have to be 100\% effective to eliminate the
transmission of a virus -- such approaches are not well-suited for use
in forecasting outcomes of potential management options. Non-parametric
approaches offer a more flexible alternative that can both more
accurately reflect the data available while also representing greater
uncertainty in areas (of state-space) where data is lacking. Ecological
research and management strategy should pay closer attention to the
opportunities and challenges nonparametric modeling can offer.

@@

We argue that while simple mechanistically motivated models may be best
for the latter case (Cuddington, Fortin, and Gerber 2013; Geritz and
Kisdi 2012), such models can be not only inaccurate but misleading in
quantitative decision making. Nonparametric models can more
realistically represent uncertainties outside of observed data while
also better capturing the dynamics in the region observed. Advances in
the theory and computational implementations of nonparametric methods
make them ripe for such applications. We use the classic problem of
optimal harvest of a marine fishery to illustrate how the nonparametric
approach of Gaussian processes can be applied.

We imagine the manager seeks to maximize some known socio-economic value
function, given their knowledge of the system. In particular, we will
focus on the case in which the manager faces a series of decisions, such
as setting a harvest quota for a marine fishery each year based on
observations of the stock size the year before.

@@

The ecological dynamics of most systems of management interest are
typically both complex and unknown {[}@refs{]}. Despite this,
quantitative policy optimization in these systems has almost exclusively
been based on simple, mechanistically motivated parametric models. There
are several reasons for this. First, limited available data almost
always precludes any accurate estimate of more complicated models.
Second, the computational demands of determining an optimal strategy in
a sequential decision-theory problem suffer from the ``curse of
dimensionality'' that makes it difficult to consider more complicated
models, states, or actions. Unfortunately, these challenges also make it
particularly difficult to reflect the true uncertainty either in the
parameters of the given models (parametric uncertainty) or in the
structure of the models themselves (structural uncertainty). Further
unknowns such as measurement uncertainty, parameter uncertainty,
unobserved states, knowledge of boundary conditions, etc. further
compound the issue. Though a hierarchical Bayesian approach provides a
natural way to address these from a statistical standpoint, formulating
reasonable parametric descriptions of each form of uncertainty is a
challenging task in itself, let alone the computational difficulties of
solving such a system. Cressie et al. (2009) provides a good account of
the successes and challenges of the approach. Applying these approaches
in the management context of sequential decision making, in which
forecasts must be obtained over a range of possible actions and updated
regularly as new information arrives makes such an approach less
feasible still.

An active literature and growing computational power over the past
several decades have only marginally improved this situation. Parametric
or structural uncertainty can be introduced only by increasing the state
space to reflect a distribution of possible parameters {[}@refs{]} or
the degree of belief in each of a set of possible models {[}@refs{]}
Frequently, the space of possible actions must then be reduced or the
algorithms adjusted by approximations to keep computations feasible.

 \textbf{\emph{This part is overstated and unnecessary}} Though machine
learning approaches have begun to appear in the ecological and
conservation literature (e.g.~in species distribution models), including
the Gaussian process based approach used here (Munch et al. 2005),
overall machine learning remains an unfamiliar and somewhat untrusted
approach for many ecologists. One potential barrier to their adoption is
the absence of a framework for applying machine learning approaches to
resource management problems.

\textbf{\emph{transition sentences!?}}

Traditional approaches to optimal control (Pontryagin's principle,
stochastic dynamic programming) rely on knowledge of the state equation,
usually described by a simple parametric model. Here we illustrate how a
stochastic dynamic programming algorithm can alternatively be driven by
the predictions from a Gaussian process -- a machine learning
approximation to the state dynamics.

\textbf{\emph{Okay, perhaps that's novel, but it's pretty trivial. Isn't
it obvious to everyone that it's trivial?}}

\textbf{\emph{paraphrase advantages of machine learning}}

Management goals / decision-theoretic approaches need accurate
prediction over relevant (short?) timescales more than reasonable
long-term mechanisms. Machine-learning approaches may offer the benefit
of the hierarchical Bayesian approach without the practical and
computational limitations of their parametric kin. Non-parametric models
are flexible enough to take maximum advantage of the data available,
while being appropriately ambiguous about the dynamics of a system in
regions of parameter space that have been poorly or never sampled.

@@

@@

Given parameters for the function $f$ and probability distribution $Z$,
along with a given economic model determining the price/profit
$\Pi(X_t, h_t)$ realized in a given year given a choice of harvest $h_t$
and observed stock $X_t$. This problem can be solved exactly for
discretized values of stock $X$ and policy $h$ using stochastic dynamic
programming (SDP) (Mangel and Clark 1985). Problems of this sort
underpin much marine fisheries management today.

A crux of this approach is correctly specifying the functional form of
$f$, along with its parameters. The standard approach {[}@refs{]} uses
one of a handful of common parametric models representing the
stock-recruitment relationship, usually after estimating the model
parameters from any available existing data. Though modifications of
this approach mentioned in the introduction can permit additional
sources of uncertainty such as measurement error in the stock
assessment, implementation errors in the harvest policy, (Clark and
Kirkwood 1986; Roughgarden and Smith 1996; Sethi et al. 2005)
uncertainty in parameters (Mangel and Clark 1985; Schapaugh and Tyre
2013) or model structure (B. K. Williams 2001; Athanassoglou and
Xepapadeas 2012), we focus on the simplest example to illustrate the
comparison, where only a single form of uncertainty (the growth shocks
$Z_t$) is present in the underlying model, making additional modes
unnecessary. \textbf{\emph{Add comment / apology on single species
dynamics?}}

We compare this approach to our alternative that uses a Gaussian Process
(GP) in place of the a given stock recruitment curve. We briefly revisit
a description of the Gaussian Process approach that has appeared in
recent literature before describing how we take the important step of
adapting the GP to the decision theory problem that can still be solved
using SDP.

@@

The use of Gaussian process (GP) regression (or ``kreging'' in the
geospatial literature) to formulate a predictive model is relatively new
in the context of modeling dynamical systems (Kocijan et al. 2005) and
introduced in the ecological modeling and fisheries management by Munch
et al. (2005). An accessible and thorough introduction to the
formulation and use of GPs can be found in C. E. Rasmussen and Williams
(2006).

The essence of the GP approach can be captured in the following thought
experiment: An exhaustive parametric approach to the challenge of
structural uncertainty might proceed by writing down all possible
functional forms for the underlying dynamical system with all possible
parameter values for each form, and then consider searching over this
huge space to select the most likely model and parameters; or using a
Bayesian approach, assign priors to each of these possible models and
infer the posterior distribution of possible models. The GP approach can
be thought of as a computationally efficient approximation to this
approach. GPs represent a large class of models that can be though of as
capturing or reasonably approximating the set of models in this
collection. By modeling at the level of the process, rather than the
level of parametric equation, we can more concisely capture the possible
behavior of these curves. In place of a parametric model of the
dynamical system, the GP approach postulates a prior distribution of
(n-dimensional) curves that can be though of as approximations to a
range of possible (parametric) models that might describe the data. The
GP allows us to consider a set of possible curves simultaneously.\\
\textbf{\emph{Figure 1 include curves drawn from the posterior
density?}}

\textbf{\emph{Do we need more specifics on Gaussian process as an
approximation to parametric models? Discussion of Gaussian process vs
other machine learning / forecasting approaches that have less clear
statistical foundations? If so, does this all belong in the
discussion?}}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{for(h in h_grid)}
  \NormalTok{F_h =}\StringTok{ }\NormalTok{for(x_j in grid)}
          \NormalTok{for(i in }\DecValTok{1}\NormalTok{:N) }
            \KeywordTok{dnorm}\NormalTok{(x_j, mu[i]-h, V[i])}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
 \NormalTok{V1 <-}\StringTok{ }\KeywordTok{sapply}\NormalTok{(}\DecValTok{1}\NormalTok{:}\KeywordTok{length}\NormalTok{(h_grid), function(h)\{}
      \NormalTok{delta *}\StringTok{ }\NormalTok{F[[h]] %*%}\StringTok{ }\NormalTok{V +}\StringTok{  }\KeywordTok{profit}\NormalTok{(x_grid, h_grid[h]) }
    \NormalTok{\})}
    \CommentTok{# find havest, h that gives the maximum value}
    \NormalTok{out <-}\StringTok{ }\KeywordTok{sapply}\NormalTok{(}\DecValTok{1}\NormalTok{:gridsize, function(j)\{}
      \NormalTok{value <-}\StringTok{ }\KeywordTok{max}\NormalTok{(V1[j,], }\DataTypeTok{na.rm =} \NormalTok{T) }\CommentTok{# each col is a diff h, max over these}
      \NormalTok{index <-}\StringTok{ }\KeywordTok{which.max}\NormalTok{(V1[j,])  }\CommentTok{# store index so we can recover h's }
      \KeywordTok{c}\NormalTok{(value, index) }\CommentTok{# returns both profit value & index of optimal h.  }
    \NormalTok{\})}
    \CommentTok{# Sets V[t+1] = max_h V[t] at each possible state value, x}
    \NormalTok{V <-}\StringTok{ }\NormalTok{out[}\DecValTok{1}\NormalTok{,]                        }\CommentTok{# The new value-to-go}
    \NormalTok{D[,OptTime-time}\DecValTok{+1}\NormalTok{] <-}\StringTok{ }\NormalTok{out[}\DecValTok{2}\NormalTok{,]       }\CommentTok{# The index positions}
\end{Highlighting}
\end{Shaded}

@@

\subsection{Example in a bistable system}

Concerns over the potential for tipping points in ecological dynamics
(Scheffer et al. 2001) highlight the dangers of uncertainty in
ecological management and pose a substantial challenge to existing
decision-theoretic approaches (Brozović and Schlenker 2011). Because
intervention is often too late after a tipping point has been crossed
(but see Hughes et al. (2013)), management is most often concerned with
avoiding potentially catostrophic tipping points before any data is
available at or following a transition that would more clearly reveal
these regime shift dynamics (e.g. Bestelmeyer et al. 2012).

Where $Z_t$ is multiplicative noise function with mean 1, representing
stochastic growth. We will consider log-normal noise with shape
parameter $\sigma_g$. We start with an example in which the parameters
are $r =$ 2, $K =$ 10, $C =$ 5, and $\sigma_g =$ 0.05.

Both parametric and nonparametric approaches will require training data
on which to base their model of the process. We generate the training
data under the model described in Eq 1 for 35 time steps, under a known
but not necessarily optimal sequence of harvest intensities, $h_t$. For
simplicity we imagine a fishery that started from zero harvest pressure
and has been gradually increasing the harvest.

@@

We estimate two parametric models from the data using a maximum
likelihood approach. The first model is structurally identical to the
true model (Eq 1), differing only in that it's parameters are estimated
from the observed data rather than given. The alternative model is the
Ricker model, which is structurally similar and commonly used in for
such data.

The maximum likelihood estimates (MLE) will assume that the process
noise $Z_t$ is log-normal, as in the simulation itself. In reality the
noise structure might differ from this assumption, making this a
best-case scenario for the MLE approach.\\The GP must also make an
assumption about the noise structure -- in this case we assume additive
noise, though multiplicative noise would be more accurate. This will
illustrate the flexibility of the GP in fitting different noise
structures even without knowledge of the correct assumptions.

The GP estimate is Bayesian approach which requires specification of
prior distributions for the hyper-parameters (see Appendix for details,
or @Rassmussen1996, Munch et al. 2005 for background), as given in Table
1. Then given the data, both the GP and the MLE parametric models can be
estimated as described above, as we implement in R in the code provided
in supplementary materials.

@@

Figure 1 shows the inferred Gaussian Process compared to the true and
the MLE parametric models, predicting the relationship between observed
stock size (x-axis) to the stock size following recruitment the
following year. The gray shading shows the uncertainty estimated by the
GP. Unlike the MLE estimates, the GP uncertainty changes depending on
where the data is available, allowing for a much greater level of
uncertainty at low stock sizes. While it would be natural (and straight
forward) to condition the GP on passing through the origin (0,0) (see
appendix), the estimate shown here is based only on the observed data.
The observed data from which each model is estimated is also shown. The
observations come from only a limited region of state space
corresponding to unharvested or weakly harvested system. No observations
occur at the theoretical optimum (7.5 units, see Figure 2) or near the
tipping point (5 units).

\begin{figure}[htbp]
\centering
\includegraphics{figure/gp_plot.png}
\caption{Graph of the inferred Gaussian process compared to the true
process and maximum-likelihood estimated process. Graph shows the
expected value for the function $f$ under each model. Two standard
deviations from the estimated Gaussian process covariance (grey shading)
is also shown. The training data is also shown as black points.}
\end{figure}

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{color}
\usepackage{fancyvrb}
\DefineShortVerb[commandchars=\\\{\}]{\|}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\newenvironment{Shaded}{}{}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.56,0.13,0.00}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{{#1}}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textit{{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{{#1}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{{#1}}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.02,0.16,0.49}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{{#1}}}}
\newcommand{\NormalTok}[1]{{#1}}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Avoiding tipping points in the management of ecological systems: a non-parametric Bayesian approach},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

\begin{document}
\begin{frontmatter}
  \title{Avoiding tipping points in the management of ecological systems: a
         non-parametric Bayesian approach}
  \author[cstar]{Carl Boettiger\corref{cor1}}
  \cortext[cor1]{Corresponding author}
  \ead{cboettig@ucsc.edu}
  \author[cstar]{Marc Mangel}
  \author[noaa]{Stephan B. Munch}
  \address[cstar]{Center for Stock Assessment Research, Department of Applied Math and Statistics, University of California, Mail Stop SOE-2, Santa Cruz, CA 95064, USA}
  \address[noaa]{Southwest Fisheries Science Center, National Oceanic and Atmospheric Administration, 110 Shaffer Road, Santa Cruz, CA 95060, USA}
 \end{frontmatter}

@@

Most management recommendations from the ecological literature are based
on (or at least motivated by) parametric models. Though in principle
these models reflect mechanistic underpinnings of biological
interactions involved, in practice real mechanisms are often unknown and
the true dynamics too complex to be captured by simple models. While
simple mechanistic models can nevertheless provide important insights
into possible dynamics -- for instance, demonstrating that a vaccine
does not have to be 100\% effective to eliminate the transmission of a
virus -- such approaches are not well-suited for use in forecasting
outcomes of potential management options. Non-parametric approaches
offer a more flexible alternative that can both more accurately reflect
the data available while also representing greater uncertainty in areas
(of state-space) where data is lacking. Ecological research and
management strategy should pay closer attention to the opportunities and
challenges nonparametric modeling can offer.

@@

Though simple mechanistically motivated models may be best for the
latter case (Cuddington, Fortin, and Gerber 2013; Geritz and Kisdi
2012), such models can be not only inaccurate but misleading in
quantitative decision making. This arises for two reasons:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  We do not know what the correct parameter values are for the models.
\item
  We do not know what the correct models are for ecological systems.
\end{enumerate}

The complexity of ecological interactions and a lack of data contribute
greatly to both of the problems. This concern is particularly acute in
case of ecological tipping points (Barnosky et al. 2012; Scheffer et al.
2001), which arise from feedbacks common in ecological systems and can
lead to a sudden catastrophic transition to an undesirable state. In
general we do not know where such tipping points are unless we have
already observed the transition, in which case it is frequently too late
to respond. Though there may exist early warning signals for certain
kinds of these transitions that are driven by slow changes (Scheffer et
al. 2009), we do not know when, where, or how to apply them to the
decision making context more generally (Boettiger and Hastings 2013).

In addition to facing the potentially grave consequences of such
transitions, if they do exist in a system of interest, this generally
means that we lack data in the region near and beyond a possible tipping
point. As a consequence, our models are least accurate just where we
need them to be most accurate.\\Unfortunately, parametric modeling
approaches are not well-suited to this scenario, in which good data is
avialble over only part of the relevant state-space. Necessary
assumptions about model structure in any parametric approach can lead
such models to perform very poorly, even when accounting for parameter
uncertainty.

Non-parametric models can more realistically represent uncertainties
outside of observed data while also better capturing the dynamics in the
region observed. Advances in the theory and computational
implementations of nonparametric methods make them ripe for such
applications. We use the classic problem of optimal harvest of a marine
fishery to illustrate how the nonparametric approach of Gaussian
processes can be applied. We will compare Bayesian implementations of
both nonparametric and parametric models, which best allow us to capture
the uncertainty of model estimation in either case and permits a more
natural comparison between approaches.

\subsubsection{Terminology: ``Non-parametric''}

The term non-parametric to describe modeling approaches such as Gaussian
processes is a common source of confusion, notwithstanding the fact that
the model is still specified by parameters. Some literature has
introduced the term ``semi-parametric'' in tacit acknowledgment of this,
which no doubt only contributes to the confusion. The problem is further
exacerbated by the several meanings assigned to the term in statistics:
(a) that the method does not assume some particular probability
distribution (e.g.~a non-parametric bootstrap) or (b) that the method
does not assume a fixed structure to the model. Our use is that of (b),
in contrast to the classical approaches that do. The difference is
clearest by way of example.

@@

The ecological dynamics of most systems of management interest are
typically both complex and unknown. Despite this, quantitative policy
optimization in these systems has almost exclusively been based on
simple, mechanistically motivated parametric models. There are several
reasons for this. First, limited available data almost always precludes
any accurate estimate of more complicated models. Second, the
computational demands of determining an optimal strategy in a sequential
decision-theory problem suffer from the ``curse of dimensionality'' that
makes it difficult to consider more complicated models, states, or
actions. Unfortunately, these challenges also make it particularly
difficult to reflect the true uncertainty either in the parameters of
the given models (parametric uncertainty) or in the structure of the
models themselves (structural uncertainty). Further unknowns such as
measurement uncertainty, parameter uncertainty, unobserved states,
knowledge of boundary conditions, etc. further compound the issue.
Though a hierarchical Bayesian approach provides a natural way to
address these from a statistical standpoint, formulating reasonable
parametric descriptions of each form of uncertainty is a challenging
task in itself, let alone the computational difficulties of solving such
a system. Cressie et al. (2009) provides a good account of the successes
and challenges of the approach. Applying these approaches in the
management context of sequential decision making, in which forecasts
must be obtained over a range of possible actions and updated regularly
as new information arrives makes such an approach less feasible still.

An active literature and growing computational power over the past
several decades have only marginally improved this situation. Parametric
or structural uncertainty can be introduced only by increasing the state
space to reflect a distribution of possible parameters or the degree of
belief in each of a set of possible models Frequently, the space of
possible actions must then be reduced or the algorithms adjusted by
approximations to keep computations feasible.

Traditional approaches to optimal control (Pontryagin's principle,
stochastic dynamic programming) rely on knowledge of the state equation,
usually described by a simple parametric model. Here we illustrate how a
stochastic dynamic programming algorithm can alternatively be driven by
the predictions from a Gaussian process.

@@

\subsection{Underlying Model}

Concerns over the potential for tipping points in ecological dynamics
(Scheffer et al. 2001) highlight the dangers of uncertainty in
ecological management and pose a substantial challenge to existing
decision-theoretic approaches (Brozović and Schlenker 2011). Because
intervention is often too late after a tipping point has been crossed
(but see Hughes et al. (2013)), management is most often concerned with
avoiding potentially catastrophic tipping points before any data is
available at or following a transition that would more clearly reveal
these regime shift dynamics (e.g. Bestelmeyer et al. 2012).

Where $Z_t$ is multiplicative noise function with mean 1, representing
stochastic growth. We will consider log-normal noise with shape
parameter $\sigma_g$. We start with an example in which the parameters
are $r =2$, $K=8$, $C=5$ and $\sigma_g = 0.1$.

Both parametric and nonparametric approaches will require training data
on which to base their model of the process. We generate the training
data under the model described in Eq 1 for $T_{\textrm{obs}}=40$ time
steps, under a known but not necessarily optimal sequence of harvest
intensities, $h_t$. For simplicity we imagine a fishery that started
from zero harvest pressure and has been gradually increasing the
harvest.

@@

Given parameters for the function $f$ and probability distribution $Z$,
along with a given economic model determining the price/profit
$\Pi(X_t, h_t)$ realized in a given year given a choice of harvest $h_t$
and observed stock $X_t$. This problem can be solved exactly for
discretized values of stock $X$ and policy $h$ using stochastic dynamic
programming (SDP) (Mangel 1985). Problems of this sort underpin much
marine fisheries management today.

A crux of this approach is correctly specifying the functional form of
$f$, along with its parameters. The standard approach uses one of a
handful of common parametric models representing the stock-recruitment
relationship, usually after estimating the model parameters from any
available existing data. Uncertainty in the parameter estimates can be
estimated and integrated over to determine the optimal policy under
under uncertainty (Mangel 1985; Schapaugh and Tyre 2013). Uncertainty in
the model structure itself can only be addressed in this approach by
hypothesizing alternative model structures, and then performing some
model choice or model averaging (B. K. Williams 2001; Athanassoglou and
Xepapadeas 2012). For simplicity of the comparison we consider only the
case of a structurally correct and structurally incorrect model, and
estimate Bayesian posterior distributions of the parameters. We will
also compare to a simpler maximum likelihood estimate of the parameters
that ignores the issue of parameter uncertainty.

We compare these parametric approaches to our alternative that uses a
Gaussian Process (GP) in place of a given stock recruitment curve. We
briefly revisit a description of the Gaussian Process approach that has
appeared in recent literature before describing how we take the
important step of adapting the GP to the decision theory problem that
can still be solved using SDP.

@@

The use of Gaussian process (GP) regression (or ``kreging'' in the
geospatial literature) to formulate a predictive model is relatively new
in the context of modeling dynamical systems (Kocijan et al. 2005) and
introduced in the ecological modeling and fisheries management by Munch
et al. (2005). An accessible and thorough introduction to the
formulation and use of GPs can be found in Rasmussen and Williams
(2006).

The essence of the GP approach can be captured in the following thought
experiment: An exhaustive parametric approach to the challenge of
structural uncertainty might proceed by writing down all possible
functional forms for the underlying dynamical system with all possible
parameter values for each form, and then consider searching over this
huge space to select the most likely model and parameters; or using a
Bayesian approach, assign priors to each of these possible models and
infer the posterior distribution of possible models. The GP approach can
be thought of as a computationally efficient approximation to this
approach. GPs represent a large class of models that can be though of as
capturing or reasonably approximating the set of models in this
collection. By modeling at the level of the process, rather than the
level of parametric equation, we can more concisely capture the possible
behavior of these curves. In place of a parametric model of the
dynamical system, the GP approach postulates a prior distribution of
(n-dimensional) curves that can be though of as approximations to a
range of possible (parametric) models that might describe the data. The
GP allows us to consider a set of possible curves simultaneously.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{for(h in h_grid)}
  \NormalTok{F_h = for(x_j in grid)}
          \NormalTok{for(i in }\DecValTok{1}\NormalTok{:N) }
            \KeywordTok{dnorm}\NormalTok{(x_j, mu[i]-h, V[i])}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
 \NormalTok{V1 <- }\KeywordTok{sapply}\NormalTok{(}\DecValTok{1}\NormalTok{:}\KeywordTok{length}\NormalTok{(h_grid), function(h)\{}
      \NormalTok{delta * F[[h]] %*% V +  }\KeywordTok{profit}\NormalTok{(x_grid, h_grid[h]) }
    \NormalTok{\})}
    \CommentTok{# find havest, h that gives the maximum value}
    \NormalTok{out <- }\KeywordTok{sapply}\NormalTok{(}\DecValTok{1}\NormalTok{:gridsize, function(j)\{}
      \NormalTok{value <- }\KeywordTok{max}\NormalTok{(V1[j,], }\DataTypeTok{na.rm =} \NormalTok{T) }\CommentTok{# each col is a diff h, max over these}
      \NormalTok{index <- }\KeywordTok{which.max}\NormalTok{(V1[j,])  }\CommentTok{# store index so we can recover h's }
      \KeywordTok{c}\NormalTok{(value, index) }\CommentTok{# returns both profit value & index of optimal h.  }
    \NormalTok{\})}
    \CommentTok{# Sets V[t+1] = max_h V[t] at each possible state value, x}
    \NormalTok{V <- out[}\DecValTok{1}\NormalTok{,]                        }\CommentTok{# The new value-to-go}
    \NormalTok{D[,OptTime-time}\DecValTok{+1}\NormalTok{] <- out[}\DecValTok{2}\NormalTok{,]       }\CommentTok{# The index positions}
\end{Highlighting}
\end{Shaded}

@@

@@

We estimate posterior distributions for two parametric models: one using
the structurally correct model as given in Eq (1), which we refer to as
the ``Parametric Bayes'' model, and another using the familiar Ricker
model, using a Gibbs sampler as described (with source code) in the
appendix). In addition we estimate the parameters of the structurally
correct model by maximum likelihood.

The posterior distribution for the hyper-parameters of the Gaussian
process model are estimated by Metropolis-Hastings algorithm, again with
details and code provided in the Appendix. Rasmussen and Williams (2006)
provides an excellent general introduction to Gaussian Processes and
Munch et al. (2005) first discusses their application in this context.

@@

\begin{figure}[htbp]
\centering
\includegraphics{Figure1.pdf}
\caption{Mean dynamics predicted by each model.}
\end{figure}

Figure 1 shows the mean inferred state space dynamics of each model
relative to the true model used to generate the data, predicting the
relationship between observed stock size (x-axis) to the stock size
after recruitment the following year. All models except the MLE model
estimate a distribution around the means shown here, and all models
estimate a level of process noise, which is independent of the state
value (x). Note that in contrast to the other models shown, the mean
Gaussian process corresponds to a distribution of curves - as indicated
by the gray band - which itself has a mean shown in black. Note that
this mean GP is thus more certain of the dynamics in the region where
data is available then where it is not.

While it would be natural (and straight forward) to condition the GP on
passing through the origin (0,0) (see appendix), the estimate shown here
is based only on the observed data. The observed data from which each
model is estimated is also shown. The observations come from only a
limited region of state space corresponding to unharvested or weakly
harvested system. No observations occur at the theoretical optimum or
near the tipping point.

\begin{figure}[htbp]
\centering
\includegraphics{Figure2.pdf}
\caption{Steady-state optimal policies by each model.}
\end{figure}

The resulting optimal management strategy based on each of the inferred
models is shown in Figure 2, against the optimal strategy given the true
underlying dynamics. Policies are shown in terms of target escapement,
$S_t$. Under models such as this a constant escapement policy is
expected to be optimal (Reed 1979), whereby population levels below a
certain size $S$ are unharvested, while above that size the harvest
strategy aims to return the population to $S$, resulting in the
hockey-stick shaped policies shown.

\begin{figure}[htbp]
\centering
\includegraphics{Figure3.pdf}
\caption{Managing the simulated system over 100 replicate instances
under each policy}
\end{figure}

The consequences of managing 100 replicate realizations of the simulated
fishery under each of the policies estimated is shown in Figure 3. As
expected from the policy curves, the structurally correct model
under-harvests, leaving the stock to vary around it's un-fished optimum.
The structurally incorrect Ricker model over-harvests the population
passed the tipping point consistently, resulting in the immediate crash
of the stock and thus derives minimal profits.
